An algorithm for finding the global extremum of a partially defined function\thanks{This work was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-08-29 2023/MMST/samplepaper.tex
add more references

Often such information is represented in the form of suggestion that the objective function $\phi$ satisfies the Lipschitz condition with a priori unknown constant $L$.

This suggestion can be interpreted (in the relation to the applied problems) as the reflection of a limited power generating the changes in the simulated system. At that, the objective function defined in the form of a ``\textit{black box}'' type functions as a rule is a non differentiable one, and each computation of its value in some point within a feasible domain may require considerable computational resources.

A number of efficient deterministic algorithms for solving the Lipschitz global optimization problems have been developed ~\cite{Jones2009,Birect2020,Sergeyev2017}. The comparisons conducted have shown the deterministic algorithms to overcome (in various criteria) the widespread nature-inspired algorithms ~\cite{Liberti2005,Sergeyev2018,Sovrasov2019} .

The present work continues the development of one of the efficient deterministic algorithms for solving the Lipschitz global optimization problems, namely the of the information-statistical global search algorithm ~\cite{indexMethod,strongin1978,Strongin2000}. This algorithm suggests the solving of a multidimensional problem by means of its reduction to an equivalent one-dimensional optimization problem using the Peano-Hilbert space-filling curves \cite{Sergeyev2013}.

Note that the use of complex mathematical models is natural in the optimization of complex real world objects that, as a consequence, essentially increases the computation costs for searching the optimum. In last decades, the experts in the fields of optimization and parallel computations have proposed many methods for reducing the computation costs an speed-up of the algorithms related to solving the appearing optimization problems \cite{Kvasov2013}  as well as to the numerical analysis of the original models \cite{Dongarra2022,Duwe2020}.

However, a principally new problem in applications, namely the numerical instability of the models investigated in some (unknown a priori) regions in the ranges of the parameters’ variations recently became relevant. In such regions, one cannot perform the numerical modeling and compute the objective function values correctly. This phenomenon can be interpreted either as the presence of some hidden (unknown a priori) constraints or as a partial computability of the objective function in the search domain. In such a statement, the optimization problem becomes more complex since the region of feasible parameters’ combinations is not defined in advance.

@@

\subsection{Description of the Global Search Algorithm}

Within the framework of the proposed approach, the information-statistical theory of global search is used for solving the problems (\ref{eq2}). The main idea of the algorithm is the use of the accumulated information to determine the next interval, within which it is most probable to find the absolute minimum. The next trial point in this interval corresponds to the mathematical expectation of the minimum location.

Let us give more detailed description of the algorithm.

\subsubsection{Global search algorithm (GSA)} At every iteration of the global search a \textit{trial} is executed. We will refer to computing the objective function value  $\phi (x)$ from (\ref{eq2}) as a trial. The first trial will be executed in the middle point $x^1 \in (0,1)$. The choice of the point $x^{k+1}, k \leq 1$ of the next $(k+1)^\text{th}$ trial is performed on the base of the following rules.

@@

A large number of excess trials after getting into the non-computability region due to a large value of the characteristic obtained according to (\ref{eq13}) is an essential draw-back of this approach.

@@

\subsubsection{The fourth approach.} To restore the objective function values in the points for the intervals with two non-computable points as the averages of the nearest computable points taking into account the coordinates, i.e.
\begin{equation}\label{eq17} 
\tilde{z}(x)=z'+ \text{sign}(z''-z') \frac {|z''-z'| \cdot \| x-x' \|}{\| x''-x' \|},
\end{equation}
where $z'$ and $z''$ are the objective function values in the computable points $x'$ and $x''$  nearest to the interval.
The rest of the approach is identical to the third one. Formula (\ref{eq14}) in this case will be applied when the non-computable region is located only near the boundary point.
\begin{figure}
\includegraphics[width=\textwidth]{fig4.png}
\caption{Imputation of non-computable values by averages of the neighboring points. Approach \#4} \label{fig4}
\end{figure}
\begin{figure}
\includegraphics[width=\textwidth]{fig5.png}
\caption{Imputation of non-computable values by averages of the minorant and the majorant. Approach \#5} \label{fig5}
\end{figure}

\subsubsection{The fifth approach.} To utilize the search information and the suggestions on the function form most completely, one can restore the objective function values in the points taking onto account the H{\"o}lder condition.
\paragraph{Step 1} Find the end points of the crossing of majorant and minorant
\begin{equation}\label{eq18} 
\hat{x}',\hat{x}''=\frac {x'+x''}{2}\pm \text{sign}(z''-z')\cdot \frac {1}{2r} \cdot {\left(\frac {z''-z'}{\mu}\right)}^N,
\end{equation}
\paragraph{Step 2} Restore the objective function values according to the rule
\begin{equation}\label{eq19} 
\tilde{z}(x)=
  \begin{cases}
    z', & {\quad x \leq \hat{x}',}\\
    \frac {z'+z''}{2}- \frac {r \mu}{2} (\| x - x' \| + \| x - x'' \|), & {\quad \hat{x}' < x < \hat{x}'',}\\
    z'',  & {\quad x \geq \hat{x}''.}
  \end{cases}
\end{equation}
The value $\tilde{z}(x)$ can be interpreted as an average value of the minorant and the majorant for a function satisfying the H{\"o}lder condition. Let us give the description of the modification of the global search algorithm in the case of partially defined function based on this approach.

@@

        \paragraph{Case 1.} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the original rules of the global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1,}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1,}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1.}
  \end{cases}
\end{equation}
        \paragraph{Case 2.}   For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0,}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1.}
  \end{cases}
\end{equation}
        \paragraph{Case 3.} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1,}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0.}
  \end{cases}
\end{equation}
        \paragraph{Case 4.} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, \; v(x_{i-1})=v(x_i)=0.
\end{equation}
        \paragraph{Case 5.} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the interval with computable boundary points that has the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{fig6.png}
\caption{Examples of problems \#\# 19, 25, and 35 from the first series of experiments} \label{fig6}
\end{figure}

>>>>

Often such information is represented in the form of suggestion that the objective function $\phi$ satisfies the Lipschitz condition with a priori unknown constant $L$.
This suggestion can be interpreted (in the relation to the applied problems) as the reflection of a limited power generating the changes in the simulated system. At that, the objective function defined in the form of a ``\textit{black box}'' type functions as a rule is a non differentiable one, and each computation of its value in some point within a feasible domain may require considerable computational resources.

A number of efficient deterministic algorithms for solving the Lipschitz global optimization problems have been developed ~\cite{Grishagin2018,Jones2021,Birect2020,PaulaviciusZilinskas2014,Sergeyev2017}. The comparisons conducted have shown the deterministic algorithms to overcome (in various criteria) the widespread nature-inspired algorithms ~\cite{Liberti2005,Sergeyev2018}.

The present work continues the development of one of the efficient deterministic algorithms for solving the Lipschitz global optimization problems, namely the of the information-statistical global search algorithm ~\cite{Strongin2000,Sergeyev2013}. This algorithm suggests the solving of a multidimensional problem by means of its reduction to an equivalent one-dimensional optimization problem using the Peano-Hilbert space-filling curves.

Note that the use of complex mathematical models is natural in the optimization of complex real world objects that, as a consequence, essentially increases the computation costs for searching the optimum. In last decades, the experts in the fields of optimization and parallel computations have proposed many methods for reducing the computation costs an speed-up of the algorithms related to solving the appearing optimization problems \cite{Kvasov2013,Sergeyev2020} as well as to the numerical analysis of the original models \cite{Dongarra2022,Duwe2020}.

However, a principally new problem in applications, namely the numerical instability of the models investigated in some (unknown a priori) regions in the ranges of the parameters’ variations recently became relevant. In such regions, one cannot perform the numerical modeling and compute the objective function values correctly. This phenomenon can be interpreted either as the presence of some hidden (unknown a priori) constraints \cite{Stripinis2021} or as a partial computability of the objective function in the search domain \cite{Candelieri2019,Sergeyev2003,Strongin2020}. In such a statement, the optimization problem becomes more complex since the region of feasible parameters' combinations is not defined in advance.

@@

\subsection{Global Search Algorithm}

%Within the framework of the proposed approach, the information-statistical theory of global search is used for solving the problems (\ref{eq2}). The main idea of the algorithm is the use of the accumulated information to determine the next interval, within which it is most probable to find the absolute minimum. The next trial point in this interval corresponds to the mathematical expectation of the minimum location.

Let us give more detailed description of the \textbf{global search algorithm} (GSA) for solving the problem (\ref{eq2}).

At every iteration of the global search a \textit{trial} is executed. We will refer to computing the objective function value  $\phi (y(x))$ from (\ref{eq2}) as a trial. The first trial will be executed in the middle point $x^1 \in (0,1)$. The choice of the point $x^{k+1}, k \leq 1$ of the next $(k+1)^\text{th}$ trial is performed on the base of the following rules.

@@

A large number of excess trials after getting into the non-computability region due to a large value of the characteristic obtained according to (\ref{eq13}) is an essential drawback of this approach.

@@

\subsubsection{The fourth approach.} To restore the objective function values in the points for the intervals with two non-computable points as the averages of the nearest computable points taking into account the coordinates, i.e.
\begin{equation}\label{eq17} 
%\tilde{z}(x)=z'+ \text{sign}(z''-z') \frac {|z''-z'| \cdot \| x-x' \|}{\| x''-x' \|},
\tilde{z}(x)=z'+ \text{sign}(z''-z') \frac {|z''-z'| \cdot \rho(x,x')}{\rho(x',x'') },
\end{equation}
where $z'$ and $z''$ are the objective function values in the computable points $x'$ and $x''$  nearest to the interval and $\rho(x_1,x_2) =  |x_1 - x_2|^{1/N}$.
The rest of the approach is identical to the third one. Formula (\ref{eq14}) in this case will be applied when the non-computable region is located only near the boundary point.
\begin{figure}
\includegraphics[width=\textwidth]{fig4.png}
\caption{Imputation of non-computable values by averages of the neighboring points. Approach \#4} \label{fig4}
\end{figure}
\begin{figure}
\includegraphics[width=\textwidth]{fig5.png}
\caption{Imputation of non-computable values by averages of the minorant and the majorant. Approach \#5} \label{fig5}
\end{figure}

\subsubsection{The fifth approach.} To utilize the search information and the suggestions on the function form most completely, one can restore the objective function values in the points taking onto account the H{\"o}lder condition.
\paragraph{Step 1} Find the end points of the crossing of majorant and minorant
\begin{equation}\label{eq18} 
\hat{x}',\hat{x}''=\frac {x'+x''}{2}\pm \text{sign}(z''-z')\cdot \frac {1}{2r} \cdot {\left(\frac {z''-z'}{\mu}\right)}^N,
\end{equation}
\paragraph{Step 2} Restore the objective function values according to the rule
\begin{equation}\label{eq19} 
\tilde{z}(x)=
  \begin{cases}
    z', & {\quad x \leq \hat{x}',}\\
    %\frac {z'+z''}{2}- \frac {r \mu}{2} (\| x - x' \| + \| x - x'' \|), & {\quad \hat{x}' < x < \hat{x}'',}\\
    \frac {z'+z''}{2}- \frac {r \mu}{2} (\rho(x,x')  + \rho(x,x'')), & {\quad \hat{x}' < x < \hat{x}'',}\\
    z'',  & {\quad x \geq \hat{x}'',}
  \end{cases}
\end{equation}
where $\rho(x_1,x_2) =  |x_1 - x_2|^{1/N}$.
The value $\tilde{z}(x)$ can be interpreted as an average value of the minorant and the majorant for a function satisfying the H{\"o}lder condition. Let us give the description of the modification of the global search algorithm in the case of partially defined function based on this approach.

@@

        \paragraph{Case 1.} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the original rules of the global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1,}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1,}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1.}
  \end{cases}
\end{equation}
        \paragraph{Case 2.}   For the intervals with \textbf{one non-computable point and one boundary point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0,}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1.}
  \end{cases}
\end{equation}
        \paragraph{Case 3.} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1,}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0.}
  \end{cases}
\end{equation}
        \paragraph{Case 4.} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, \; v(x_{i-1})=v(x_i)=0.
\end{equation}
        \paragraph{Case 5.} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the interval with computable boundary points that has the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{fig6.png}
\caption{Examples of problems No. 19, 25, and 35 from the first series of experiments} \label{fig6}
\end{figure}
